---
title: "ACH's Base R cheatsheet"
output: html_notebook
---

## Commands
**open shortcuts pop-up** - `alt + shift + k` <br>
**clear console** - `ctrl + L` <br>
**move line** - `alt + up/down arrow` <br>
**duplicate code** - `ctrl + shift + D` <br>
**Go to file/function popup** - `ctrl + .` (period) <br>
**navigate panes = ctrl + nums** - `ctrl + 1` & `ctrl + 2` <br>
**navigate to terminal** - `alt + shift + T`
**create a notebook** - File > New File > R Notebook <br>
**pipe %>% operator** - `ctrl + shift + m` <br>

```{r}
# An 'R Notebook' is different from just 'R markdown...'.
# 
# File > New File > 'R Notebook'
# is the one you want. It's similar to a Jupyter Notebook in python
# At the top you will see:

#---
#title: "R Notebook"
#output: html_notebook
#---



```


## Packages
**install libraries** (aka packages) `install.packages("packageName")`

**load library** `library(packageName)`

**detailed info about current session for reproducible results**
`sessionInfo()`

**list loaded packages** `(.packages())`

**check for conflicts among packages** `conflicts(detail=TRUE)`

**check for package dependencies**
`tools::package_dependencies("packageName", db = installed.packages())`

**check if package is a dependency for other packages**
`tools::package_dependencies("packageName", db = installed.packages(), reverse = TRUE)`


## Load Data

#### Tidyverse - readr package

You can load data by filename only if your data is in the same directory as your R Notebook/script.<br> Otherwise you will need either (file = relative path) or (file = absolute path).<br>
**csv** `data <- read_csv(file = 'name.csv', colnames = TRUE/FALSE)`

**.Rdata files** `load("fileName.Rdata")`

## Look at Data
**get structure of data** - `str(data)` <br>
**view data in new tab** - `View()` <br>


## Vectors (aka lists or arrays)

#### Base R: vectors (lists/arrays)
```{r lists}
# create a vector
earnings <- c(100, -150, 250, -50, 350)

# assign column names
names(earnings) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# show new vector with col names
earnings
```

#### Base R: combining vectors with math
When you add vectors, each element gets added by element position. So 1st + 1st, 2nd + 2nd, 3rd + 3rd. This is a very different result from calling the sum function. Adding vectors is very different from getting the sum of vectors.
```{r vector math}
# adding vectors: each element gets added by position
c(1, 2, 3) + c(4, 5, 6)

x <- c(2, 4, 6)
y <- c(3, 6, 9)

# add vectors x and y
x + y

# subtract vectors x and y
x - y

# multiply vectors x and y
x * y

# divide vectors x and y
x / y

# find the average of vector x
mean(x)

# find the average of the sum
mean(x+y)
```

#### Base R: - `sum()`
With a single argument, this function adds up the elements within a vector. With multiple arguments, the `sum()` function adds up all the elements within each vector. This is different from adding two different vectors element-wise: 1st + 1st, 2nd + 2nd, etc. like we did in the examples above with `x` and `y`.
```{r sum()}
a <- c(5, 10, 15)
sum(a)

b <- c(30, 40, 50)
sum(a, b)
```

#### Base R: select elements in a vector - by index
In R, vectors are indexed at 1, unlike C or Python languages where arrays or lists are 0-indexed.
```{r element by index}
# create a vector of days of the week
days <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday",
          "Friday", "Saturday")

# select the first element
days[1]

# select elements 2 through 4
days[2:4]

# select only the 2nd and 6th days of the week
days[c(2, 6)]
```
#### Base R: select elements in a vector - by name.
```{r element by name}
# output earnings vector with named elements
earnings

# select an element by it's col name
earnings["Wednesday"]

# select only Tuesday & Thursday earnings by name
earnings[c("Tuesday", "Thursday")]

# 'days' vector only containts M-F. What happens when we try to select all the
# earnings based on the named vector?
earnings[days]
```

#### Base R: Logical Operators - select elements in a vector - by comparison
You can use comparison operators on entire vectors.
`<`, `<=`, `>`, `>=`, `==`, `!=`, `x&y`, `x|y`, `!x` (Not x), `isTRUE(x)` test if X is TRUE.
```{r element by comparison}
# show vector 'a'
a

# compare which elements in vector 'a' less than 12
a < 12

# are any elements in vector 'a' not 10?
a != 10

# selecting elements in a vector based on comparison operators
z <- c(10, 20, 30, 40, 50)
z[z > 30]
```


## Tidyverse

#### `summarize()`

You can use summarize with the following functions:

`mean()` <br>
`sum()` <br>
`median()` <br>
`min()` <br>
`max()` <br>

