---
title: "Working with Categorical Variables -- aka Factors"
output: html_notebook
---

## What's a factor?
The term factor refers to a statistical data type used to store categorical variables. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a limited number of categories. A continuous variable, on the other hand, can correspond to an infinite number of values. A good example of a categorical variable is sex. In many circumstances you can limit the sex categories to "Male" or "Female". 

#### Create a factor
To create factors in R, you make use of the function `factor()`. First thing that you have to do is create a vector that contains all the observations that belong to a limited number of categories.
```{r observations vector}
# example: sex_vector contains the sex of 5 different individuals
sex_vector <- c("Male","Female","Female","Male","Male")
```

It is clear that there are two categories, or in R-terms 'factor levels', "Male" and "Female".

```{r create factor vector}
# 'factor()' will encode the vector as a factor:
factor_sex_vector <- factor(sex_vector)

# show the factor vector
factor_sex_vector
```
