---
title: "Working with Categorical Variables -- aka Factors"
output: html_notebook
---

## What's a factor?
The term factor refers to a statistical data type used to store categorical variables. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a limited number of categories. A continuous variable, on the other hand, can correspond to an infinite number of values. A good example of a categorical variable is sex. In many circumstances you can limit the sex categories to "Male" or "Female". 

### Create a factor
To create factors in R, you make use of the function `factor()`. First thing that you have to do is create a vector that contains all the observations that belong to a limited number of categories.
```{r observations vector}
# example: sex_vector contains the sex of 5 different individuals
sex_vector <- c("Male", "Female", "Female", "Male", "Male")
```

It is clear that there are two categories, or in R-terms 'factor levels', "Male" and "Female".

```{r create factor vector}
# 'factor()' will encode the vector as a factor:
factor_sex_vector <- factor(sex_vector)

# show the factor vector
factor_sex_vector
```

### Types of categorical variables: nominal, ordinal.

A nominal variable is a categorical variable without an implied order. This means that it is impossible to say 'one is worth more than the other'. Think "nominal = name". Consider animals, `elephant`, `giraffe`, `hippo`.

In contrast, ordinal variables do have a natural ordering. Consider rankings `good`, `better`, `best`. We know the order of these categories. Still, we don't know *how much* better one category is than the other.


### Changing the Categories (factor levels)
When you first get a data set, you will often notice that it contains factors with specific factor levels. However, sometimes you will want to change the names of these levels for clarity or other reasons. R allows you to do this with the function `levels()`.

* `levels(factor_vector) <- c("name1", "name2",...)`

**Nominal vector**
```{r make levels}
# record survey data in vector
survey <- c("M", "F", "F", "M", "M")
# notice there are no levels yet
survey

# turn survey vector into a factor vector
factor_survey <- factor(survey)
# show the vector, notice the levels?
factor_survey
```


**Watch out**: the order with which you assign the levels is important. Notice above, the levels outputs `"F" "M"`. If you don't specify the levels of the factor when creating the vector, R will automatically assign them alphabetically. To correctly map "F" to "Female" and "M" to "Male", the levels should be set to c("Female", "Male"), in this order.
```{r change levels}
levels(factor_survey)

# change the levels/categories of the factor vector
levels(factor_survey) <- c('Female', 'Male')

# show the factor vector with renamed levels
factor_survey
```

`summary()` on a factor vector will give you a quick count of factors and quantity in each.
```{r summarizing-a-factor}
summary(factor_survey)
```

Sometimes you will also deal with factors that do have a natural ordering between its categories. If this is the case, we have to make sure that we pass this information to R.

**Ordinal vector**
```{r}
# create a vector that with ordered values
speed_vector <- c('medium', 'slow', 'slow', 'medium', 'fast')

# convert speed vector to ordered factor vector
factor_speed_ordered <- factor(speed_vector,
                               ordered = TRUE,
                               levels = c('slow', 'medium', 'fast'))

# show the vector and summarize the vector
factor_speed_ordered
summary(factor_speed_ordered)
```

With an ordinal vector, you can use comparison operators to analyze different elements.

### Compare elements of an ordinal vector
```{r compare-ordinal-elements}
# show the ordinal vector
factor_speed_ordered

# compare elements in an ordinal vector
runner3 <- factor_speed_ordered[3]
runner4 <- factor_speed_ordered[4]
runner3 > runner4
```

