---
title: "Conditionals"
output: html_notebook
---

# If and Else statements

## If statement
Assign a value to a variable, then write an if-test. If the condition is true, do what's in the braces.
```{r}
# if-statement: condition true
x <- -3

if (x < 0){
    print("x is negative")
}

# if-statement: condition false
if (x > 0){
    print("x is positive")
}
```

## Else statement

`else` statement does not need an explicit conditional, but it must be used with an `if-statement`.
```{r}
# set initial value
x <- -3

if(x > 0){
    print("x is positive or zero")
} else {
    print("x is negative")
}
```

## Else-if statement

The `else-if` statement comes in between the `if` and the `else` statement. It's important that the `else if` keywords comes on the same line as the closing bracket of the previous part of the control construct!

```{r}
# set the initial value
x <- -3

if (x < 0){
    print("x is negative")
} else if (x == 0){
    print("x is zero")
} else {
    print("x is positive")
}
```

## Caution

If your conditions are not mutually exclusive you will run into problems. As soon as R encounters an expression that evaluates to TRUE it will execute the code and ignore any code in the rest of the statement.

```{r}
# a problem occurs when our conditions are not mutually exclusive
# example: divisible by 2 and 3

x <- 6

if (x %% 2 == 0){
    print("x is divisible by 2")
} else if (x %% 3 == 0){
    print("x is divisible by 3")
} else {
    print("x is not divisble by 2 or 3")
}
```

Notice x is actually divisible by both 2 and 3, but it never got to the 2nd conditional.