---
title: "Conditionals"
output: html_notebook
---

# If and Else statements

## If statement
Assign a value to a variable, then write an if-test. If the condition is true, do what's in the braces.
```{r}
# if-statement: condition true
x <- -3

if (x < 0){
    print("x is negative")
}

# if-statement: condition false
if (x > 0){
    print("x is positive")
}
```

## Else statement

`else` statement does not need an explicit conditional, but it must be used with an `if-statement`.
```{r}
# set initial value
x <- -3

if(x > 0){
    print("x is positive or zero")
} else {
    print("x is negative")
}
```

## Else-if statement

The `else-if` statement comes in between the `if` and the `else` statement.
```{r}
# set the initial value
x <- -3

if (x < 0){
    print("x is negative")
} else if (x == 0){
    print("x is zero")
} else {
    print("x is positive")
}
```

## Caution
If your conditions are not mutually exclusive you will run into problems. As soon as R encounters an expression that evaluates to TRUE.