---
title: "Control Flow & Conditionals"
output: html_notebook
---

## Relational Operators
Relational operators are for comparisons. They let us see how one R object relates to another.

* TRUE: 1
* FALSE: 0
* Equality `==`
* Inequality `!=`
* Greater than `>`
* Less than `<`
* Greater than or equal to `>=`
* Less than or equal to `<=`


```{r equality}
# compare strings
"compare" == "strings"

# compare numbers
5 == 10

# check if hello is not goodbye
"hello" != "goodbye"

```

## Greater or Less than - alphabetical
For character strings, the greater-than less-than operators check the first letter of the string. For string comparison, R determines the greater than relationship based on alphabetical order. 
```{r}
# since 'H' comes after 'G' in the alphabet, "Hello" is greater than "Goodbye"
"Hello" > "Goodbye"

# compare similar phrases
"raining" <= "raining dogs"

# compare capitalization
"useR" > "user"
```

## Relational Operators with Vectors
Say you want to compare your profile views - github v facebook
```{r}
# create two profile views vectors for the example
github <- c(5, 25, 10, 15, 0, 30, 20)
fb <- c(50, 250, 10, 7, 35, 100, 25)

# check which days the github views exceeded 10
github > 10

# check when number of github views are less than or equal to fb views
github <= fb
```

## Relational Operators with Matrices
R's ability to deal with different data structures for comparisons does not stop at vectors. Matrices and relational operators also work together seamlessly.
```{r}
# create a 2x7 matrix using the github and fb vectors
views <- matrix(c(github, fb), nrow = 2, byrow = TRUE)
views

# create row and column name vectors
profiles <- c("github", "fb")
days <- c("M", "T", "W", "R", "F", "Sat", "Sun")

# add names to matrix
rownames(views) <- profiles
colnames(views) <- days

# show new matrix
views
```

Comparisons in a matrix
```{r}
# when does views = 13?
views == 13

# when are views <= 15?
views <= 15
```

## Logical Operators

* AND operator `&` - returns TRUE only if both logical values are true
* OR operator `|` - returns TRUE if either logical value is true
* NOT operator `!`

### AND operator
```{r}
TRUE & TRUE

TRUE & FALSE

FALSE & TRUE

FALSE & FALSE
```

### AND operator on numerical expressions
```{r}
# set variable x for example
x <- 10

# if both logicals are true, this evaluates to TRUE
x > 5 & x < 15

# if either epxression is false, this evaluates to FALSE
x > 5 & x < 9

```


### OR operator
```{r}
TRUE | TRUE

TRUE | FALSE

FALSE | TRUE

FALSE | FALSE
```

### OR operator on numerical expressions
```{r}
# set variable 'y' for example
y <- 5

# if either expression is true, it evaluates to TRUE
y < 6 | y > 10

# if neither expression is true, it evaluates to FALSE
y < 3 | y > 10
```

## NOT operator
```{r}
# if something is NOT true, it evaluates to FALSE
!TRUE

# if something is NOT false, it evaluates to TRUE
!FALSE

# there are multiple ways to evaluate the same expressions
!(x < 5)
x >= 5
```

The NOT operator can be really handy when used with other R functions such as `is.numeric()` which checks to see if an R objec is numeric.
```{r}
is.numeric(5)

!is.numeric(5)

is.numeric("hello")

!is.numeric("hello")
```

## Logical Operators with Vectors & Matrices

Just like relational operators, logical operators perform the operations in an **element-wise** manner.
```{r}
# AND
c(TRUE, TRUE, FALSE) & c(TRUE, FALSE, FALSE)

# OR
c(TRUE, TRUE, FALSE) | c(TRUE, FALSE, FALSE)

# NOT
!c(TRUE, TRUE, FALSE)
```

## Logical Operators - single or double `&` v `&&`, `|` v `||`

The results of your test will be different depending on if you use a single or double logical operator. The biggest difference is when you use them on vectors. The double ampersand operation only examines the first element of each vector

```{r single-double-AND}
# single logical operators v. double logical operators

# AND operators - &, &&
# single
c(TRUE, TRUE, FALSE) & c(TRUE, FALSE, FALSE)

# double - only examines the first element in each vector
c(TRUE, TRUE, FALSE) && c(TRUE, FALSE, FALSE)

c(FALSE, TRUE, FALSE) && c(TRUE, FALSE, FALSE)
```

```{r single-double-OR}
# OR operators - |, ||
# single
c(TRUE, TRUE, FALSE) | c(TRUE, FALSE, FALSE)

# double
c(FALSE, TRUE, FALSE) || c(TRUE, FALSE, FALSE)
```


