---
title: "Working with Data Frames"
output: html_notebook
---

## What are data frames? How are they different from matrices?
**Matrix** - all the elements in a matrix should be of the same type.

**Data frame** - can have elements that are of different types. columns = variables. rows = observations.

R has a built-in package called 'datasets'. You can view a list of datasets in a new tab in Rstudio by calling `data()` function.
```{r included-data}
# look at the structure of the 'mtcars' data set
str(mtcars)
```

Notice that every variable is type `numeric` in this data set.

### Display portions of data
```{r show-data}
# show the first 6 rows
head(mtcars)

# show the last 6 rows
tail(mtcars)
```

## Build your own data set
```{r build-data}
# define your vectors
name <- c("Mercury", "Venus", "Earth",
          "Mars", "Jupiter", "Saturn",
          "Uranus", "Neptune")

type <- c("Terrestrial planet",
          "Terrestrial planet",
          "Terrestrial planet",
          "Terrestrial planet",
          "Gas giant",
          "Gas giant",
          "Gas giant",
          "Gas giant")

diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# create a Base R data frame (different from tidyverse)
planets_df <- data.frame(name, type, diameter, rotation, rings)
planets_df
```

## Select elements of a data frame
`df_name[row, col]`
```{r select-elements}
# Print out diameter of Mercury (row 1, column 3)
planets_df[1, 3]

# Print out data for Mars (entire fourth row)
planets_df[4, ]
```
