---
title: "Working with Data Frames"
output: html_notebook
---

## What are data frames? How are they different from matrices?
**Matrix** - all the elements in a matrix should be of the same type.

**Data frame** - can have elements that are of different types. columns = variables. rows = observations.

R has a built-in package called 'datasets'. You can view a list of datasets in a new tab in Rstudio by calling `data()` function.
```{r included-data}
# look at the structure of the 'mtcars' data set
str(mtcars)
```

Notice that every variable is type `numeric` in this data set.

## Display portions of data
```{r show-data}
# show the first 6 rows
head(mtcars)

# show the last 6 rows
tail(mtcars)
```

## Build your own data set
```{r build-data}
# define your vectors
name <- c("Mercury", "Venus", "Earth",
          "Mars", "Jupiter", "Saturn",
          "Uranus", "Neptune")

type <- c("Terrestrial planet",
          "Terrestrial planet",
          "Terrestrial planet",
          "Terrestrial planet",
          "Gas giant",
          "Gas giant",
          "Gas giant",
          "Gas giant")

diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# create a Base R data frame (different from tidyverse)
planets_df <- data.frame(name, type, diameter, rotation, rings)
planets_df
```

## Select elements of a data frame
### `df_name[row, col]`
```{r select-elements}
# Print out diameter of Mercury (row 1, column 3)
planets_df[1, 3]
```

```{r}
# Print out data for Mars (entire fourth row)
planets_df[4, ]
```

## Select by row/col number
### `planets_df[1:3, 2]`.
Instead of using numerics to select elements of a data frame, you can also use the variable names to select columns of a data frame. Suppose you want to select the first three elements of the `type` column. A possible disadvantage of this approach is that you have to know (or look up) the column number of type, which gets hard if you have a lot of variables.

## Select by row/col name
###  `planets_df[1:3, "type"]`
It is often easier to just make use of the variable name.

```{r}
head(planets_df)
```

```{r}
# Select & print out the first 5 values in the "diameter" column of planets_df
planets_df[1:5, 3]
```

## Select entire column
You will often want to select an entire column, namely one specific variable from a data frame.
```{r select-column}
# select column by number
planets_df[,3]

# select column by name
planets_df[,"diameter"]

# shortcut using '$'
diameter_vector <- planets_df$diameter
```

## Use columns to get values from other columns
Say you wan to get the names of the planets with rings. You're going to use the rings column to get only the planets where rings = TRUE.
```{r}
planets_df

# isolate the rings column into a vector
rings_vector <- planets_df$rings

# if planet has rings, get the name of the planet
planets_df[rings_vector, 1:5]
```

## `subset()` is a shortcut for the same thing
### `subset(my_df, subset = some_condition)`
```{r subset-method}
# get the planet names whose diameter is smaller than Earth (Earth = 1)
subset(planets_df, subset = diameter < 1)
```

